---
title: "Pool 4 Lower Exploration"
author: "Amber Lee"
date: "6/3/2021"
output: pdf_document
---

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(corrplot)
library(RColorBrewer)
library(reticulate)
```

pictures, geographical information about the pool
some data information

* for example, casey only has main channel and side channel data. (what are the strata that we have?)

* statistical summary of 4 variables: `TP`, `CHLcal`, `SS`, `TURB`

* number of observations per strata

* future directions / ideas

* an interesting paper or article from USGS website relating to pool 4 lower

* missing data

* extreme outliers

* `TP` and `TN` may have some missing values

(next friday: learn to pose own questions within an ecosystem framework, explain the project in our own words)

```{r}
veg19 <- read.csv(file = "../pool data/ltrm_vegsrs_data_lat_long.csv")
water20 <- read.csv(file = "../pool data/ltrm_water_data_lat_long.csv")
# lter use data through 2020 that has lat/lng
```

## Vegetation data

```{r}
veg19 %>% 
  filter(POOL == "04",
         str_detect(MSTRATUM, "-L")) %>%
  select(DATE, DETRITUS, SBSTRATE, VEG_S, VEG_RF, 
         VEG_E, VEG_A, VEG_NRF, VEG_U,
         COV_NRF, COV_RF, COV_E,
         RAKE1, RAKE2, RAKE3, RAKE4, RAKE5, RAKE6) %>% head()

veg_4L <- veg19 %>% 
  filter(POOL == "04",
         str_detect(MSTRATUM, "-L")) %>%
  mutate(DATE = mdy(DATE))
```

Rake distribution

```{r}
veg_4L %>% 
  mutate(year = year(DATE)) %>%
  select(year, RAKE1, RAKE2, RAKE3, RAKE4, RAKE5, RAKE6) %>%
  pivot_longer(!year, names_to = "subsample_rake",
               values_to = "level") %>%
  group_by(year) %>%
  summarize(mean_level = mean(level),
            sd_level = sd(level)) %>%
  ggplot(aes(x = year, y = mean_level, 
             fill = sd_level)) +
  geom_point()
  
  # 
  # ggplot(aes(x = RAKE1)) +
  # geom_histogram()
```

Water depth

```{r}
veg_4L %>% 
  mutate(year = year(DATE)) %>%
  select(year, starts_with("DEPTH")) %>%
  pivot_longer(!year, names_to = "subsample",
               values_to = "depth") %>%
  ggplot(aes(x = year, y = depth, group = year)) +
  geom_boxplot()
  
  # 
  # ggplot(aes(x = RAKE1)) +
  # geom_histogram()
```

plant density

```{r}
veg_4L %>% 
  mutate(year = year(DATE)) %>%
  select(year, starts_with("DENSITY")) %>%
  pivot_longer(!year, names_to = "subsample",
               values_to = "density") %>%
  ggplot(aes(x = year, y = density, group = year)) +
  stat_summary(
    mapping = aes(x = year, y = density),
    fun.min = function(z) { quantile(z,0.25) },
    fun.max = function(z) { quantile(z,0.75) },
    fun = mean)
```

## Water data

```{r}

water20 %>%
  filter(FLDNUM == "1", !is.na(STRATUM)) %>%
  mutate(`Pool 4 Regions` = case_when(LONGITUDE <= -92.33 ~ "Lower",
                                      LONGITUDE >= -92.11 ~ "Upper",
                                      TRUE ~ "A lake"),
         Stratum = case_when(STRATUM == 1 ~ "Main channel",
                             STRATUM == 2 ~ "Side channel",
                             STRATUM == 3 ~ "Backwater area",
                             STRATUM == 4 ~ "Lake Pepin",
                             STRATUM == 5 ~ "Impounded",
                             STRATUM == 6 ~ "Isolated",
                             STRATUM == 7 ~ "New Terrestrial"),
         DATE = mdy(DATE),
         year = year(DATE)) %>%
  ggplot(aes(x = LATITUDE, y = LONGITUDE)) +
  geom_point(aes(color = Stratum), alpha = 0.2) +
  scale_color_manual(values = brewer.pal(9,'Blues')[3:10]) +
  geom_abline(intercept = -92.11, slope = 0) +
  geom_abline(intercept = -92.33, slope = 0) +
  ggtitle("Pool 4", 
          subtitle = "Black lines indicate how we split upper and lower") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  scale_x_continuous(limits = c(44.3, 44.65)) +
  scale_y_continuous(limits = c(-92.65, -91.8))

ggsave("splitting_pool4.png")


#3000 out of 204k observations removed

```

```{r}

water_var <- c('TN','TP','TEMP','DO','TURB',
               'COND','VEL','SS','WDP','CHLcal','SECCHI')

# filter for pool 4 lower
water_4L <- water20 %>% 
  filter(FLDNUM == "1") %>%
  mutate(DATE = mdy(DATE),
         year = year(DATE),
         is_lower = case_when(LONGITUDE <= -92.33 ~ TRUE,
                              TRUE ~ FALSE),
         decade = case_when(year <= 2000 ~ "1993-2000",
                            year >= 2001 & year <= 2014 ~ "2001-2014",
                            year >= 2014 ~ "2014-2020"),
         Stratum = case_when(STRATUM == 1 ~ "Main channel",
                             STRATUM == 2 ~ "Side channel",
                             STRATUM == 3 ~ "Backwater area",
                             STRATUM == 4 ~ "Lake Pepin",
                             STRATUM == 5 ~ "Impounded",
                             STRATUM == 6 ~ "Isolated",
                             STRATUM == 7 ~ "New Terrestrial")) %>%
  filter(is_lower & STRATUM != 4)


```

```{r}

ggplot(water_4L, aes(x = LATITUDE, y = LONGITUDE)) +
  geom_point(aes(color = Stratum), alpha = 0.75, size = 2) +
  scale_color_manual(values = brewer.pal(7,'Blues')[2:7]) +
  ggtitle("Pool 4L Stratum", 
          subtitle = "Black lines indicate how we split upper and lower") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  scale_y_continuous(limits = c(-92.625, -92.4))

ggsave("stratum_pool4L.png")
  
```

### Sampling distribution

```{r}

ggplot(water_4L) +
  geom_histogram(aes(x = year, fill = Stratum)) +
  facet_wrap(~ Stratum) +
  scale_fill_manual(values = brewer.pal(7,'Blues')[2:7]) +
  ggtitle("Sampling distribution of strata per year",
          subtitle = "Pool 4 Lower")

ggsave("sampling_distribution_year_strata.png")

```


### Missing values

* number of missing values per variable per year

* Missingness by decade (zooming out)

```{r}

water_4L <- bind_cols(water_4L, missing = rowSums(is.na(water_4L)))

# data.frame(Variables = names(water_4L), Missing = colSums(is.na(water_4L))) %>%
#   filter(Variables %in% water_var) %>%
#   ggplot(aes(x = Variables, y = Missing)) +
#   geom_bar(stat = "identity")

water_4L %>%
  group_by(year) %>%
  summarise(across(water_var, ~ sum(is.na(.x)))) %>%
  pivot_longer(water_var, names_to = "Variables") %>%
  ggplot(aes(x = year, y = value)) +
  geom_point(aes(color = Variables)) +
  geom_line(aes(color = Variables)) +
  ggtitle("Number of missing values per year",
          subtitle = "Of selected variables (color coded) in Pool 4 Lower")

ggsave("missing_count_selected_pool4L.png")

water_4L %>%
  select(year, all_of(c("TN", "TURB", "SS", "CHLcal"))) %>%
  group_by(year) %>%
  summarise(across(c("TN", "TURB", "SS", "CHLcal"), ~ sum(is.na(.x))) ) %>%
  pivot_longer(c("TN", "TURB", "SS", "CHLcal"), names_to = "Variables") %>%
  ggplot(aes(x = year, y = value)) +
  geom_point(aes(color = Variables)) +
  geom_line(aes(color = Variables)) +
  facet_wrap(~ Variables) +
  ggtitle("Number of missing values per year",
          subtitle = "Of selected variables (color coded) in Pool 4 Lower")

ggsave("missing_count_4_pool4L.png")

ggplot(water_4L, aes(x = decade, y = missing)) +
  geom_violin(aes(group = decade, fill = decade)) +
  ggtitle("Density of count of missing observations per decade",
          subtitle = "Pool 4 Lower") +
  scale_fill_manual(values = brewer.pal(7,'Blues')[3:6]) +
  theme(legend.position = "none")
  
  
ggsave("missing_violin_pool4L.png", height = 4, width = 6)

```

### Continuous variable box plots
```{r}

plotter_box_by_year <- function(var_str, data, facet_bool){
  
  # facet_bool gives if you should facet by STRATUM type
  
  title <- paste(var_str, "boxplot_by_year", "pool4L", sep = "_")
  
  if (facet_bool){title <- paste(title, "_facet")}
  
  title <- paste(title, "png", sep = ".")
  
  if (facet_bool){
    data %>%
      filter(!is.na(!!sym(var_str))) %>%
      ggplot(aes(x = year, y = !!sym(var_str), group = year)) +
      geom_boxplot() + 
      facet_wrap(~ Stratum, nrow = 2)
  } else {
    data %>%
      filter(!is.na(!!sym(var_str))) %>%
      ggplot(aes(x = year, y = !!sym(var_str), group = year)) +
      geom_boxplot()
  }
  
  ggsave(title)

}

sapply(c("TP", "TN", "TURB", "SS", "CHLcal"), plotter_box_by_year, water_4L, F)
sapply(c("TP", "TN", "TURB", "SS", "CHLcal"), plotter_box_by_year, water_4L, T)

plotter_box_by_year("VEL", water_4L, F)
ggsave("VEL_boxplot_by_year_pool4L.png")
plotter_box_by_year("VEL", water_4L, T)
ggsave("VEL_boxplot_by_year_pool4L_facet.png")

# SS, turbidity, and secchi all three measure water quality

# velocity graph is new

# dissolved oxygen is variable by the time of day

# temperature is important per season

# habitat class for these water quality data is all NA
# unique(water_4L$HABCLASS)

unique(water_4L$STRATUM)

table(water_4L$STRATUM)

```

Note: stratum types
1 = Main channel
2 = Side channel
3 = Backwater area contiguous to the main channel
4 = Lake Pepin or Swan Lake


* total nitrogen is variable by year, but within year, not too variable

* velocity has increased, water volume is greater. increase in velocity reflects collection in data? collect

### Temperature

```{r}

# tried to use location code for the seasons but recorded differently...

seasonalwater <- water_4L %>%
  mutate(month = month(DATE),
         Season = case_when(month >= 2 & month <= 4 ~ "Spring",
                             month >= 5 & month <= 7 ~ "Summer",
                             month >= 8 & month <= 10 ~ "Fall",
                             TRUE ~ "Winter"))

seasonalwater$Season <- factor(seasonalwater$Season, 
                              levels = c("Spring", "Summer",
                                         "Fall", "Winter"))

ggplot(seasonalwater, aes(x = year, y = TEMP, group = year)) +
  geom_boxplot(aes(color = Season)) +
  facet_wrap(~ Season) +
  ggtitle("Temperature recordings by season") +
  theme(legend.position = "none") +
  scale_color_manual(values = brewer.pal(6, "YlGnBu")[3:6])

ggsave("seasonal_temperature_pool4L.png")

ggplot(seasonalwater, aes(x = year, y = DO, group = year)) +
  geom_boxplot(aes(color = Season)) +
  facet_wrap(~ Season) +
  ggtitle("Dissolved Oxygen recordings by season") +
  theme(legend.position = "none") +
  scale_color_manual(values = brewer.pal(6, "YlGnBu")[3:6])

seasonalwater %>% filter(CHLcal <= 100) %>%
  ggplot(aes(x = year, y = CHLcal, group = year)) +
  geom_boxplot(aes(color = Season)) +
  facet_wrap(~ Season) +
  ggtitle("Chlorophyll A recordings by season",
          subtitle = "With outlier measurements about 100 removed") +
  theme(legend.position = "none") +
  scale_color_manual(values = brewer.pal(6, "YlGnBu")[3:6])

ggsave("seasonal_chlorophyllA_pool4L.png")

# 60 observations removed (out of 6k)

```

### Water quality

```{r}

waterqual_4L <- water_4L %>% 
                     select(SS, TURB, SECCHI)

waterqual_4L <- waterqual_4L %>%
  bind_cols(missing = rowSums(is.na(waterqual_4L))) %>%
  filter(missing == 0) %>% # remove missing
  select(-missing)

ggplot(data = water_4L, aes(x = SS, y = TURB)) +
  geom_point(aes(color = log(SECCHI)))

corrplot(round(cor(waterqual_4L ), digits = 4))

ggsave("waterquality_corrplot.png")


```


[website](umesc.usgs.gov/ltrm-home.html)

## Euclidean distance metric

https://www.rdocumentation.org/packages/pracma/versions/1.9.9/topics/haversine

```{r}
library(pracma) # haversine formula
```

```{r}
loc_df <- water20 %>%
  select(LATITUDE, LONGITUDE, TP) %>%
  distinct() %>%
  head(10)
```

Here, latitude is given in north
Longitude is given in west (hence negative)
```{r}


getLatLng <- function(n_col, df) {
  
  latlng <- c(df[n_col,1],
              df[n_col,2])
  
  return(latlng)
  
}

getDist <- function(n_iterating_location, df, location){
  
  iterating_location <- getLatLng(n_iterating_location, df)
  
  return(haversine(iterating_location, location))
  
}

getColDist <- function(n_col, df){
  
  location <- getLatLng(n_col, df)
  
  col_name <- paste(n_col, "_loc", sep = "")
  
  col <- data.frame(eval(parse(text = col_name)) = sapply(1:dim(df), getDist, df, location))
  
  # iterate through each row of the df to generate ONE row
  return(col)

}

getColDist(1, loc_df)

bind_cols(lapply(1:dim(loc_df), getColDist, loc_df))


```

```{r}
big_loc_df <- water20 %>% 
  select(LONGITUDE, LATITUDE, TP) %>%
  head(1000) 

lapply(1:dim(big_loc_df), getColDist, big_loc_df)

```


```{r}

water20 %>% 
  select(SHEETBAR, TIME, DATE, LONGITUDE, LATITUDE) %>%
  group_by(SHEETBAR, TIME, DATE, LONGITUDE, LATITUDE) %>%
  summarize(count = n()) %>%
  filter(count > 1)

water20 %>% filter(DATE == "06/11/1993")


```

```{r}

water20 %>% filter(SHEETBAR == "-4303746")

```


```{r}

water20 %>% 
  filter(SAMPZCD != CALCZCD) %>%
  select(SHEETBAR, TIME, DATE, LONGITUDE, LATITUDE) %>%
  group_by(SHEETBAR, TIME, DATE, LONGITUDE, LATITUDE) %>%
  summarize(count = n()) %>%
  filter(count > 1)

```

```{r}
water20 %>% distinct(SHEETBAR) %>% dim()
```


SAMPLE_DEPTH_CODE                  
SAMPZCD             aa             SF,BT,MD,OT                     1988 TO PRESENT                                   
CALC_SAMPLE_DEPTH_CODE             CALCZCD             aa             SF,BT,MD,OT 
